---
title: "Eddie Ice Phenology R Notebook"
output: html_notebook
---
# Eddie Module Ice Phenology     
This is an adaption of the eddie module on ice phenology to examine regressions of ice on off and duration of ice cover from lakes.    
     
The first step is to use a demonstration of Lake Sunapee, New Hampshire, USA then the script will allow students to adapt the code to be use to examine other lakes.       
    
# Install libraries     
Install  libraries for use to examine data using tidyverse this only has to be done one time and a # can be put in front of the code to comment it out so it does not run in the future as now.   
```{r install libraries}
# install.packages("tidyverse")
# install.packages("readxl")
# install.packages("janitor")
# install.packages("lubridate")
# install.packages("broom")
```

# Load libraries    
Load the libraries each time you run a script or program   
```{r load libraries}
library(tidyverse) # loads a lot of libraries in this one to add more functionality
library(readxl) # allows you to read in excel files
library(janitor) # cleans up column names and removes empty columns if wanted
library(lubridate) # allows easy conversion of varaibles to date format
library(broom) # cleans up output for easy presentation
```
   
# Read file   
Note since they are all in different sheets you can change         
      sheet = "name of lake"     
to the name you are interested in to read in that lake       

Read in Lake Sunapee and stores in the sunapee dataframe in the environment tab.      
    Note this is slightly different than in the regular scipt <br>     
    the `../` has to be used as this file thinks the main working directory is<br> 
    where it is located so we have to jump up one level and then use /data <br> 
    
```{r read data}
sunapee.df <- read_excel("../data/lake_ice_phenology_module-.xls",  
                         # identifies the file to read in
  sheet="Sunapee", # identifies the sheet to read in
  guess_max = 1000) # allows more rows to guess the type of data
```

# Convert Dates to proper date format 
Behind this date is the number of seconds since 1970 or number of days
```{r creat dates}
sunapee.df  <- sunapee.df  %>% # use the sunapee dataframe and passes to next command
  mutate(ice_on_date = mdy(ice_on_date), 
         # converts to date using mdy as Month Day Year format 
         ice_off_date = mdy(ice_off_date), 
         # converts to date using mdy as Month Day Year format 
         ice_on_julian = yday(ice_on_date), 
         # converts to date using mdy as Month Day Year format 
         ice_off_julian = yday(ice_off_date))
         # converts to date using mdy as Month Day Year format 
```

# Create new varaible for break point analsys 
if_else tells r to evaluate a column and do things      
if true is the first option if false is the second option    
can change 1970 to another year if you want a different break point
```{r create breakpoint}
sunapee.df  <- sunapee.df %>% # pass sunapee date to next command
         mutate( # mutate changes things
           pre_post_1970 = ifelse(start_year <= 1970, "pre_1970", "post_1970") 
           # if else to make breakpoint of 1970 ^^^^, if true^^, if false^^^
           ) 
```


# Regression of ice_off_doy 
Work with lakes separately to calculate regression coefficients
The first line of code below stores a linear model in an object called sunapee.model
lm = linear model of ice_off_julian day is a function of start year using the data from sunapee.df    
   
```{r linear model}
sunapee.model <- lm(ice_off_julian ~ start_year, data = sunapee.df)
```


# Model Summary
To see the output from the model you run this below    
```{r model summary}
summary(sunapee.model)
```

## Model Output
`Call: lm(formula = ice_off_julian ~ start_year, data = sunapee.df) `    
`     `
`Residuals: # this is an estimate of the statistics of the residuals`     
`  Min      1Q  Median      3Q     Max   `  
`-27.297  -5.122   1.268   6.250  16.644  `   
     
**these are the coefficients**     
**(Intercept) = the y intercept but since years are used the intercept is not very useful**     
**start_year = the slope of the line**     
**the (>|t|) is the probability the value is not 0 = means nothing really**     
`Coefficients:     `
`              Estimate    Std. Error  t value   Pr(>|t|)     `
`  (Intercept) 294.35842     35.03231    8.402   3.82e-14 ***     `
`  start_year   -0.09269      0.01804   -5.138   8.87e-07 ***     `
`  ---`
`  Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1    ` 
`     `
`Residual standard error: 9.187 on 144 degrees of freedom `    
`Multiple R-squared:  0.1549,	Adjusted R-squared:  0.1491 `    
`F-statistic:  26.4 on 1 and 144 DF,  p-value: 8.872e-07 `    
     
**The F-statistic determines if the model is significant or not and this one is.**       
**The square root of the R-stquared gives you the regression coefficient of r **     
     
# Condensed output        
To get a reduced amount of infomation you can run the tidy version using Broom     
```{r tidy output}
tidy(summary(sunapee.model))
```

# Create a graph of the data     
```{r basic plot}
sunapee.df %>% # takes the dataframe and feeds it into the plot
  ggplot(aes(x = start_year, y = ice_off_julian)) + 
  # tells ggplot to use variables for x and y (note the x=/y= are usually not necessary)
  geom_point() +
  # adds points for individual observations
  geom_smooth(method="lm") + 
  # performs a linear model on the data with a 95% confidence interval shaded
  labs(y = "Ice off day (Julian Day)", 
       # creates the y axis label
       x = "Year of observation" ) 
       # creates the x axis label
```
     
# A prettier plot
You can save the plot to an object as well the graph can be exported using the export button in the graph window or as below       
```{r pretty plot}
sunapee.plot <- sunapee.df %>% # takes the dataframe and feeds it into the plot
  ggplot(aes(x = start_year, y = ice_off_julian)) + 
  # tells ggplot to use variables for x and y (note the x=/y= are usually not necessary)
  geom_point(size = 3, color="darkblue", fill = "red", shape = 25) + # adds points for individual observations
  geom_smooth(method="lm") + 
  # performs a linear model on the data with a 95% confidence interval shaded
  scale_x_continuous(
    limits = c(1860, 2020), 
    # creates the limits of the axis - note will remove data not in range
    breaks = seq(1860, 2020, by = 20) 
    # sets the lables on the axis in a sequence from and to with ticks by = 
  ) +
  labs(y = "Ice off day (Julian Day)", # creates the y axis label
       x = "Year of observation", # creates the x axis label
       title = "Lake Sunapee" # title
  ) + 
  theme( # this allows modification of the look of the graph with common settings
    axis.line = element_line(size = 0.4, linetype = "solid"), # color and size of axis lines
    axis.ticks = element_line(colour = "black",  size = 0.9),  # color and size of ticks
    axis.title = element_text(family = "Helvetica", size = 15, face = "bold"), #size of axis tit#les
    axis.text.x = element_text(family = "Helvetica", size = 16),  # size of x axis text overall
    axis.text.y = element_text(family = "Helvetica", size = 16),  # size of y axis text overall
    plot.title = element_text(family = "Helvetica", size = 19))  # title settings

sunapee.plot
```
# Saving the plot     
```{r save plot}
ggsave("../figures/sunapee ice off doy vs year plot.pdf", # save the plot to this file in figures
       sunapee.plot, # which plot to save - change this to change the plot saved
       device = "pdf", # can export as "tiff", "jpeg", "pdf", and others
       width = 6, # width of the plot in units
       height = 6, # height of the plot in units
       units = "in" ) # units to be used
```

# 1970 Break point analysis 
There are a few ways to do this analysis... 
## 1. The really manual method
This method creates separate dataframes for pre and post and then does the above approach

## Extract only sunapee pre 1970 data
```{r}
sunapee_pre1970.df <- sunapee.df %>% # use the sunapee data set 
  filter( # filter out some of the data to retain 
    pre_post_1970 == "pre_1970") # this filters out only pre_1970 data in the pre_post_1970 column
# all of this is saved in the sunapee_pre1970.df dataframe
```

## Extract only sunapee post 1970 data        
```{r}
sunapee_post1970.df <- sunapee.df %>% # use the sunapee data set 
  filter( # filter out some of the data to retain 
    pre_post_1970 == "post_1970") # this filters out only pre_1970 data in the pre_post_1970 column
# all of this is saved in the sunapee_pre1970.df dataframe
```

## Sunapee pre 1970 model    
```{r pre 1970 model}
sunapee_pre1970.model <- lm(ice_off_julian ~ start_year, data = sunapee_pre1970.df)
```

## Pre 1970 model summary 
To see the data from the model you run this below    
```{r summary pre 1970}
summary(sunapee_pre1970.model)
```

*Prey 1970 Model output*
Call: lm(formula = ice_off_julian ~ start_year, data = sunapee_pre1970.df)      
      
Residuals:      
  Min      1Q  Median      3Q     Max       
-26.992  -5.216   1.440   6.430  16.605       
      
Coefficients:      
  Estimate Std. Error t value Pr(>|t|)          
(Intercept) 314.04384   61.73621   5.087 1.71e-06 ***      
  start_year   -0.10310    0.03216  -3.206  0.00181 **       
  ---      
  Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1      
      
Residual standard error: 9.563 on 100 degrees of freedom      
Multiple R-squared:  0.0932,	Adjusted R-squared:  0.08413       
F-statistic: 10.28 on 1 and 100 DF,  p-value: 0.001808      
      
## Sunapee pre 1970 model
```{r post 1970 model}
sunapee_post1970.model <- lm(ice_off_julian ~ start_year, data = sunapee_post1970.df)
```

## Post 1970 model summary 
To see the data from the model you run this below
```{r summary post}
summary(sunapee_post1970.model)
```

*Model Output*
Call:      
  lm(formula = ice_off_julian ~ start_year, data = sunapee_post1970.df)      
      
Residuals:      
  Min       1Q   Median       3Q      Max       
-24.1395  -4.9908   0.5587   6.0430  15.9264       
      
Coefficients:      
  Estimate Std. Error t value Pr(>|t|)         
(Intercept) 641.81466  191.59047   3.350  0.00172 **      
  start_year   -0.26674    0.09615  -2.774  0.00823 **      
  ---      
  Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1      
      
Residual standard error: 8.099 on 42 degrees of freedom      
Multiple R-squared:  0.1549,	Adjusted R-squared:  0.1347       
F-statistic: 7.695 on 1 and 42 DF,  p-value: 0.008225      
      

## 2 The better approach     
The second way to do this without splititng the dataframe      
```{r}
sunapee_pre_post.model <- sunapee.df %>% 
  group_by(pre_post_1970) %>% # this grpups the data into the groups listed in this variable
  do(pre_post.model = lm(ice_off_julian ~ start_year, # this does the model in two steps 
                         data = .)) # tells it to use the data fred into the model above
```

## summary of tidy model
This saves the coefficients out to a new dataframe and does the summary of the model what model and the column to use of teh groups.     
```{r summary of tidy model}
sunapee_pre_post.coef <- tidy(sunapee_pre_post.model, pre_post.model) 
sunapee_pre_post.coef # prints out the dataframe
```

## Plotting pre and post 1970     
You can see the graph with both lines
```{r plot of break analysis}
sunapee.df %>% 
  ggplot(aes(start_year, ice_off_julian, 
             color=pre_post_1970, # this maps a color to the points based on the column
             shape=pre_post_1970, # this maps a symbol shape to the column
             linetype=pre_post_1970)) + # this changes the line time based on the column
  geom_point() + # add in points
  geom_smooth(method="lm") # add in a smoothed line

```









